A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2017-04-27 model.tex
new stuff on instability when dust is vertically well mixed. where to put in numerical example?

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature is lowered with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to in fact behave as if it had a
temperture \emph{increasing} away from $z=0$.

@@

{\bf
\subsection{Physical disk conditions}
physical disk conditions underwhich the above approx
(small/well-coupled dust and isothermal gas) is applicable. smaller
than mm at few tens of au. give dimensional numbers here
}

@@

\subsubsection{Order-of-magnitude estimates} 
We evaluate the above stability criteria for the equilibria described in \S\ref{eqm}. 
%However, here we permit 
%$\epsilon_0=\epsilon_0(r)$ and $\Htilde=\Htilde(r)$ to vary with
%radius. 
We assume the disk is approximately Keplerian so that $\kappa\simeq\OmK$, and that the vertical gas distribution is
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\epsilon}{\left(1+
    \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

@@

Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must
vary on an extremely short lengthscale. 
%an extremely short  
%radial length-scale in the dust-to-gas ratio. 
This is unlikely to
occur. On the other hand, if dust is well-mixed, say $\Hd\simeq
0.99\Hg$, then instability only requires 
$L_\epsilon\lesssim O(10\Hg)$, which is more realistic.

>>>>

Notice Eq. \ref{eos} resembles an ideal gas equation of state but with  
a reduced temperature $\widetilde{T} = 
T_\mathrm{ref}\left(1-\tepsilon\right)$. %sound-speed $c_{s,\mathrm{eff}} = c_s\sqrt{\left(1 -
This reduced temperature is lowered with dust-loading. 
%   \tepsilon\right)}$. 
%then the equation of state has the standard
%form $P=c_{s,\mathrm{eff}}^2\rho$. 
Since $\tepsilon$ typically decrease away from the midplane, we expect
vertically isothermal dusty disks to behave as if it had a temperture
\emph{increasing} away from $z=0$.

@@

%{\bf
\subsection{Physical disk conditions}
%physical disk conditions underwhich the above approx
%(small/well-coupled dust and isothermal gas) is applicable. smaller
%than mm at few tens of au. give dimensional numbers here
%}

The terminal velocity approximation applies to short
stopping times, $\tstop\OmK\ll 1$; while the locally isothermal
approximation applies to short cooling times, $t_\mathrm{cool}\OmK\ll
1$. In typical protoplanetary disk
models such as the Minimum Mass Solar Nebula, $\tstop$ decreases for
smaller particles and/or at smaller radii \citep{youdin11}. However,
$t_\mathrm{cool}$ is generally only small in the outer disk
\citep{lin15,malygin17}. Combining these results, we estimate that our
framework is best suited for mm (or smaller) sized particles at a few
to 10s of AU in protoplanetary disks. However, note that the
isothermal approximation may be relaxed to other fixed equations of
state (see \S\ref{gen_poly}).

@@

We expect Eq. \ref{dusty_solberg1} to be satisfied in typical
protoplanetary disks where the dust-to-gas ratio increases in the same
direction as the local pressure gradient, which is equivalent to the
gas density gradient for isothermal gas. Thus to enable instability
here we need to violate Eq. \ref{dusty_solberg2}. We explore this
below.

\subsubsection{Vertically well-mixed dust}
Here we consider the special case where the dust-to-gas ratio is
uniform in $z$. Then the LHS of Eq. \ref{dusty_solberg2} becomes
\begin{align}
-\frac{c_s^4}{(1+\tepsilon)^4}\left(\frac{\p\ln{\rhog}}{\p
    z}\right)^2\left(\frac{\p\epsilon}{\p r}\right)^2<0.
\end{align}
Thus the disk is \emph{unstable}, since there is no effective vertical 
buoyancy to stabilize vertical motions. In fact, this situation is
identical to the pure gas, adiabtic simulations of \cite{nelson13} in
which the gas entropy is vertically uniform but has a radial
gradient. There, \citeauthor{nelson13} indeed find instability.

\subsubsection{Order-of-magnitude estimates} 
In order to apply the stability criteria
Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2} to typical
protoplanetary disks we conside the equilibria described in
\S\ref{eqm}.

We assume the disk is approximately Keplerian so that $\kappa\simeq\OmK$, and that the vertical gas distribution is
Gaussian. In terms of the dust-to-gas
ratio $\epsilon$, Eq. \ref{dusty_solberg1}---\ref{dusty_solberg2}
become   
\begin{align}
  &1 + \frac{\epsilon}{\left(1+
    \epsilon\right)^2}\left(h_\mathrm{g}^2\frac{\p\ln{\rhog}}{\p\ln{r}}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} + 
  \frac{z^2}{\Htilde^2}\right)>0 \label{stability_est1},   \\ 
&1 - \frac{\epsilon
  h_\mathrm{g}^2}{\left(1+\epsilon\right)^2}
  \frac{\p\ln{\epsilon}}{\p\ln{r}}\left(-\frac{\p\ln{\rhog}}{\p\ln{r}}+\frac{\Htilde^2}{H_\mathrm{g}^2}\frac{\p\ln{\epsilon}}{\p\ln{r}}\right)
  > 0 \label{stability_est2},
\end{align}
\emph{for stability}.

@@

Taking $\epsilon\sim 0.01$,  we find that 
for thin dust layers with $H_\epsilon \simeq \Hd\ll \Hg$,  instability
requires $L_\epsilon\ll O(0.1\Hg)$, i.e. the dust-to-gas ratio must
vary on an extremely short lengthscale. 
% %an extremely short  
% %radial length-scale in the dust-to-gas ratio. 
% This is unlikely to
% occur. On the other hand, if dust is well-mixed, say $\Hd\simeq
% 0.99\Hg$, then instability only requires 
% $L_\epsilon\lesssim O(10\Hg)$, which is more realistic.
